Equalizer Programming Guide\\ {\footnotesize\sffamily\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-10-07 Developer/ProgrammingGuide/paper.tex
Programming Guide: update to v0.9
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.8, \today
}

@@

\thispagestyle{empty}
\begin{figure}[ht]
  \center
  \includegraphics[width=4cm]{images/logo.pdf}
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill
\begin{figure}[ht]
  \center
  \includegraphics[width=4cm]{images/vmml.pdf}
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.8     & Sep 28, 2007 & added head tracking, finished channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & started channel and image compositing,
                             added event handling, moved to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & added pipe, window and object manager\\
    0.5     & Sep 3,  2007 & added config and partly pipe\\
    0.4     & Aug 31, 2007 & added distributed objects\\
    0.3     & Aug 26, 2007 & added application and render client\\
    0.2     & Aug 20, 2007 & added main function\\
    0.1     & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

A prerequisite for this Programming Guide is a working \textsf{eqPly}
example. The Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
explains how to run it. \textsf{eqPly} can also be executed without a
server, which simplifies the development cycle. In this case it will be
configured to use one window.

@@

@@

\subsection{The Resource Tree}

@@

\section{The \textsf{eqPly} polygonal renderer}

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application. Its focus is not on rendering features or visual quality,
but it is used as a test bed for most of the Equalizer features.

@@

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window}
class. \fig{fUml} shows the most important Equalizer classes and how
they are subclassed in the \textsf{eqPly} example. Each of the classes
in the \textsf{eqPly} namespace will be discussed in this section.

@@

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and deserializing its data. The serialization is done in
\textsf{getInstanceData}. The serialized data is cached using a
(non-distributed) member variable, which is cleared by each setter of
distributed data. Serializing the data is a simple matter of allocating
a large enough memory buffer, and copying the data into the buffer:

{\footnotesize\begin{lstlisting}
const void* InitData::getInstanceData( uint64_t* size )
{
    *size = sizeof( StaticData ) + _filename.length() + 1;
    if( _instanceData )
        return _instanceData;

    _instanceData = new char[ *size ];

    reinterpret_cast< StaticData* >( _instanceData )[0] = _data;

    const char* string = _filename.c_str();
    memcpy( _instanceData + sizeof( StaticData ), string, _filename.length()+1);

    return _instanceData;
}
\end{lstlisting}}

After the data is no longer needed, \textsf{releaseInstanceData} is
called by Equalizer to allow the object to free the data. Since the data
is cached for further usage, the release method is not overwritten in
\textsf{InitData}.

The memory buffer returned by \textsf{getInstanceData} is transferred to
the remote node during object mapping and passed to
\textsf{applyInstanceData} for deserialization. The deserialization
routine reads the data back into its own member variables:

{\footnotesize\begin{lstlisting}
void InitData::applyInstanceData( const void* data, const uint64_t size )
{
    _data     = reinterpret_cast< const StaticData* >( data )[0];
    _filename = static_cast<const char*>( data ) + sizeof( StaticData );

    EQASSERT( _data.frameDataID != EQ_ID_INVALID );
    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}%>>

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
execution is synchronized later.

@@

The \textsf{eqPly::Config} class holds the master versions of the
initialization and frame data. Both objects are registered with the
\textsf{eq::Config}, which is the \textsf{eqNet::Session} used for
rendering. Equalizer takes care of the session setup and exit in
\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
respectively.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}

@@

\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

@@

In \textsf{eqPly}, the OpenGL context is first set up using various
\textsf{apply} convenience methods from the base Equalizer channel
class. Each of the \textsf{apply} methods uses the corresponding
\textsf{get} method(s) and then calls the approriate OpenGL
function(s). It is also possible to just query the values from Equalizer
using the \textsf{get} methods, and use them to set up the OpenGL state
appropriatly, for example by passing the parameters to the renderer used
by the application.

@@

The following contextual information has to be used in order to render
the view expected by Equalizer:
\begin{description}
\item[Buffer]: The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  seperation and anaglyphic stereo settings.
\item[Viewport]: The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum]: Frustum parameters as defined by
  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
  projection matrix. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass.
\item[Head Transformation:] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frustra in immersive rendering. It is normally used to set up the
  `view' part of the modelview matrix, before static light sources are
  defined.
\item[Range]: A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data.
\end{description}

Consequently, the first part of the \textsf{frameDraw} method in
\textsf{eqPly} look as follows:

{\footnotesize\begin{lstlisting}
void Channel::frameDraw( const uint32_t frameID )
{
    applyBuffer();
    applyViewport();

    glMatrixMode( GL_PROJECTION );
    glLoadIdentity();

    applyFrustum();

    glMatrixMode( GL_MODELVIEW );
    glLoadIdentity();
    applyHeadTransform();
\end{lstlisting}}

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/DB.pdf}
  {\caption{\small\label{fDB}Destination view of an DB compound}}
\end{wrapfigure}

After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render the frame in the same position.

Furthermore, a color is set in case the model does not contain color
information, or the color information is not to be used. In sort-last
rendering, \textsf{eqPly} renders the individual parts with different
colors to illustrate the algorithm, as shown in \fig{fDB}. The Equalizer
channel provides a method to obtain a random, but unique color for all
channels in the configuration:

{\footnotesize\begin{lstlisting}
    glLightfv( GL_LIGHT0, GL_POSITION, lightpos );

@@

Finally the model is rendered, if it could be loaded during
node initialization. Otherwise a quad is drawn in its place:

@@

Two task methods are responsible for collecting and compositing the
result image during scalable rendering, when multiple channels are
rendering for a single view. The source channels producing an
\textsf{outputFrame} use \textsf{Channel::frameReadback} to read the
pixel data from the frame buffer. The channels receiving one or multiple
\textsf{inputFrame} use \textsf{Channel::frameAssemb\-le} to assemble
the frames into the framebuffer. Equalizer takes care of the network
transport of frame buffer data between nodes, if needed.

@@

In order to provide a motivation for the design of the image compositing
API, the parallel compositing algorithm for direct send is introduced
here.

The motivation for direct send is to parallelize the costly reassembly
for database (sort-last) decomposition. With each additional source
channel, the amount of pixel data to be composited grows
linearly. Direct send distributes this workload evenly across all source
channels, and thereby keeps the compositing work per channel constant.

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
Each channel is responsible for the sort-last compositing of one 2D
tile. He receives the framebuffer pixels for his tike from all the other
channels. The size of this tile decreases linearly with the number of
source channels.

After performing the sort-last compositing, the color information is
transferred to the destination channel, similarly to an 2D (sort-first)
compound.

@@

The following operations have to be possible in order to perform this
algorithm:
\begin{itemize}
\item Selection of color and/or depth readback
\item Restricting the read back to a part of the rendered area
\item Positioning the pixel data correctly on the receiving channels
\end{itemize}

@@

\subsubsection{Frame, FrameData and Images}

@@

The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel with respect to the window.

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
For output frames, the frame data's pixel viewport (pvp) is the area of
the frame buffer to read back. It will transport the offset from the
source to the destination channel, that is, the frame data pvp for input
frames position the pixel data on the destination. This has the effect
that a partial framebuffer readback will end up in the same place in the
destination channels.

@@

The \textsf{eVolve} example is a scalable volume renderer.

@@

The \textsf{eqPly} example contains rudimentary support for head
tracking, in order to show how head tracking can be integrated with
Equalizer. Supporting a wide range of tracking devices is not within the
scope of Equalizer. Other open source and commercial implementations
cover this functionality sufficiently and can be integrated with
Equalizer.

>>>>

\title{Equalizer Programming Guide\\
  {\footnotesize\sffamily\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\\{\footnotesize Eyescale Software GmbH}
}
\date{
  \vfill
  \vfill
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vfill
  \vfill
  \textbf{DRAFT}\\[\medskipamount]
  Version 0.9 for Equalizer v0.4\\[\medskipamount]
  \today
}

@@

\thispagestyle{empty}

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9     & Oct 7, 2007  & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

It is recommended to have a compiled version of Equalizer and the
examples for this Programming Guide. The Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
should be executed once to get familar with the capabilities of the
\textsf{eqPly} example.

Compiling Equalizer is as simple as running \textsf{make} on Linux or
building the Equalizer solution on Windows. On Mac OS X, some
prerequisites have to be installed before running \textsf{make}, as
explained in \textsf{README.Darwin}.

@@

\section{Hello, World!}

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/eqHello.png}
  {\caption{\small\label{fHello}Hello, World!}}
\end{wrapfigure}

The \textsf{eqHello} example is a minimal example to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

The \textsf{eqHello} redraw function renders six rotating, colored quads
around the origin, where the viewer is usually positioned. The Equalizer
\textsf{frameDraw} meth\-od is used as a convience function to setup the
frustum and other OpenGL state. After setting up some additional
lighting parameter, \textsf{eqHello} rotates the scene and render the
quads using immediate mode:

{\footnotesize\begin{lstlisting}
void Channel::frameDraw( const uint32_t spin )
{
    // setup OpenGL State
    eq::Channel::frameDraw( spin );

    const float lightPos[] = { 0.0f, 0.0f, 1.0f, 0.0f };
    glLightfv( GL_LIGHT0, GL_POSITION, lightPos );

    const float lightAmbient[] = { 0.2f, 0.2f, 0.2f, 1.0f };
    glLightfv( GL_LIGHT0, GL_AMBIENT, lightAmbient );

    // rotate scene around the origin
    glRotatef( static_cast< float >( spin ) * 0.5f, 1.0f, 0.5f, 0.25f );

    // render six axis-aligned colored quads around the origin
    [...]
}
\end{lstlisting}}

The \textsf{eqHello} main function sets up the communication with the
server, initializes and drives the rendering. The details of this setup
are explained in \sref{sEqPly}.

@@

\subsection{\label{sResource}The Resource Tree}

@@

\section{\label{sEqPly}The eqPly polygonal renderer}

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application of medium complexity. Its focus is not on rendering features
or visual quality, but it is used as a show case for the integration of
most of the Equalizer features.

@@

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window} class which
is frequently used as a type by windowing systems. \fig{fUml} shows the
most important Equalizer classes and how they are subclassed in the
\textsf{eqPly} example.

The derived classes fall into two categories: The subclassed rendering
entities as discussed in \sref{sResource}, and two classes for
distributing data. For each of the rendering entities, its function and
typical usage will be discussed. The distributed data classes illustrate
the typical usage of static and dynamic, frame-specific data.

@@

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and deserializing its data. The serialization is done in
\textsf{getInstanceData} and deserialization in
\textsf{applyInstanceData} by streaming all member variable to or from
the provided data streams. The data transport between nodes is
implemented in the data streams and various other Equalizer classes:

{\footnotesize\begin{lstlisting}
void InitData::getInstanceData( eqNet::DataOStream& os )
{
    os << _frameDataID << _windowSystem << _useVBOs << _useShaders << _filename;
}

void InitData::applyInstanceData( eqNet::DataIStream& is )
{
    is >> _frameDataID >> _windowSystem >> _useVBOs >> _useShaders >> _filename;

    EQASSERT( _frameDataID != EQ_ID_INVALID );
    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}%>>

The data output and input streams perform no type checking on the
variables. It is the application's responsibility to match the order and
types of variables during serialization and deserialization exactly.

@@

The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
execution is synchronized later.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The \textsf{eqPly::Config} class holds the master versions of the
initialization and frame data. Both objects are registered with the
\textsf{eq::Config}, which is the \textsf{eqNet::Session} used for
rendering. Equalizer takes care of the session setup and exit in
\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
respectively.

@@

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

@@

In \textsf{eqPly}, the OpenGL context is first set up using various
\textsf{apply} convenience methods from the base Equalizer channel
class. Each of the \textsf{apply} methods uses the corresponding
\textsf{get} method(s) and then calls the appropriate OpenGL
function(s). It is also possible to just query the values from Equalizer
using the \textsf{get} methods, and use them to set up the OpenGL state
appropriatly, for example by passing the parameters to the renderer used
by the application.

@@

The contextual information has to be used in order to render the view as
expected by Equalizer. Failure to use certain information will result in
incorrect rendering for some or all configurations. The channel render
context consist of:

\begin{description}
\item[Buffer]: The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  seperation and anaglyphic stereo settings.
\item[Viewport]: The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum]: Frustum parameters as defined by
  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
  projection matrix. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass.
\item[Head Transformation]: A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frustra in immersive rendering. It is normally used to set up the
  `view' part of the modelview matrix, before static light sources are
  defined.
\item[Range]: A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data.
\end{description}

The \textsf{frameDraw} method in \textsf{eqPly} calls the
\textsf{frameDraw} method from the parent class, the Equalizer
channel. The default \textsf{frameDraw} method uses the apply
convenience functions to setup the OpenGL state for all render context
information, with the exception of the range which will be used later
during rendering:

{\footnotesize\begin{lstlisting}
void eq::Channel::frameDraw( const uint32_t frameID )
{
    applyBuffer();
    applyViewport();

    glMatrixMode( GL_PROJECTION );
    glLoadIdentity();
    applyFrustum();

    glMatrixMode( GL_MODELVIEW );
    glLoadIdentity();
    applyHeadTransform();
}

void Channel::frameDraw( const uint32_t frameID )
{
    // Setup OpenGL state
    eq::Channel::frameDraw( frameID );
\end{lstlisting}}

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/DB.pdf}
  {\caption{\small\label{fDB}Destination view of an DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

Furthermore, a white color is set in case the model does not contain
color information, or the color information is not to be used. In
sort-last rendering, \textsf{eqPly} uses a different color for each
channell to illustrate the database decomposition, as shown in
\fig{fDB}. The Equalizer channel provides a method to obtain a random,
but unique color for all channels in the configuration. This color is
determined by the server to ensure uniqueness across all channels of the
configuration:

{\footnotesize\begin{lstlisting}
    glLightfv( GL_LIGHT0, GL_POSITION, lightPosition );
    glLightfv( GL_LIGHT0, GL_AMBIENT,  lightAmbient  );
    glLightfv( GL_LIGHT0, GL_DIFFUSE,  lightDiffuse  );
    glLightfv( GL_LIGHT0, GL_SPECULAR, lightSpecular );

    glMaterialfv( GL_FRONT, GL_AMBIENT,   materialAmbient );
    glMaterialfv( GL_FRONT, GL_DIFFUSE,   materialDiffuse );
    glMaterialfv( GL_FRONT, GL_SPECULAR,  materialSpecular );
    glMateriali(  GL_FRONT, GL_SHININESS, materialShininess );

@@

Finally the model loaded by the node is rendered. If the model was not
loaded during node initialization, a quad is drawn in its place:

@@

Two task methods are responsible for collecting and compositing the
result image during scalable rendering, when multiple channels are
rendering for a single view. The source channels producing one or more
\textsf{outputFrame}s use \textsf{Channel::frameReadback} to read the
pixel data from the frame buffer. The channels receiving one or multiple
\textsf{inputFrame}s use \textsf{Channel::frameAssemb\-le} to assemble
the pixel data into the framebuffer. Equalizer takes care of the network
transport of frame buffer data between nodes, if needed.

Normally the programmer does not need to interfere with the image
compositing, or only at a high level, for example to order the input
frames or to optimize the readback. The following sections provide a
detailed description of the image compositing API in Equalizer.

@@

In order to provide a motivation for the design of the image compositing
API, the direct send parallel compositing algorithm is introduced in this
section.

The motivation for direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the naive approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives the
framebuffer pixels for his tile from all the other channels. The size of
one tile decreases linearly with the number of source channels, which
keeps the total amount of pixel data per channel constant.

After performing the sort-last compositing, the color information is
transferred to the destination channel, similarly to an 2D (sort-first)
compound. The amount of pixel data for this part of the compositing
pipeline also approaches a constant value, i.e., the full frame buffer.

@@

The following operations have to be possible in order to perform this
algorithm:
\begin{itemize}
\item Selection of frame buffer attachments: color and/or depth
\item Restricting the read back area to a part of the rendered area
\item Positioning the pixel data correctly on the receiving channels
\end{itemize}

@@

\subsubsection{Frame, Frame Data and Images}

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel with respect to the window.

For output frames, the frame data's pixel viewport (pvp) is the area of
the frame buffer to read back. It will transport the offset from the
source to the destination channel, that is, the frame data pvp for input
frames position the pixel data on the destination. This has the effect
that a partial framebuffer readback will end up in the same place in the
destination channels.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
texture-based volume rendering, where the 3D texture is intersected by
view-aligned slices. The slices are rendered back-to-front and blended
together to produce the final image, as shown in
\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
  Research Council (DFG)}.

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in this mode.

The full power of scalable volume rendering is however in DB (sort-last)
compounds, where the full volume is divided into separate bricks. Each
of the bricks is rendered the same as in the other modes. For
recomposition, the \textsf{RGBA} frame buffer data resulting from these
render passes then has to be assembled correctly.

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependend data (less data per time step and channel).

\begin{figure}[h!t]
  \subfigure[]{\label{fBlendOrtho}
    \includegraphics[width=.45\textwidth]{images/b2f_ortho.pdf}
  }\hfil
  \subfigure[]{\label{fBlendPerspective}
    \includegraphics[width=.55\textwidth]{images/b2f_perspective.pdf}
  }%
  {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
      Orthogonal and Perspective View Frustra}}
\end{figure}
For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For currect blending, the frames are ordered in
the same back-to-front order as the slices used for rendering using the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

Finding the correct assembly order for a perspective frustra is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frustra. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2. In order to compute the assembly order, \textsf{eVolve} uses
the angle between the origin$\rightarrow$slice vector and the near
plane, as shown in \fig{fBlendPerspective}. The result image naturally
looks the same as the volume rendering would when rendered n a singly
channel, as shown in \fig{fVolResult}.

@@

The \textsf{eqPly} example contains rudimentary support for head
tracking, in order to show how head tracking can be integrated with
Equalizer. Supporting a wide range of tracking devices is not within the
scope of Equalizer. Other open source and commercial implementations
cover this functionality sufficiently and can easily be integrated with
Equalizer.
